---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
# set working directory
setwd("~/Coursera FULL/5 -- Reproducible Research/RepData_PeerAssessment1")

# load used libraries
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)


options(scipen=999) # no scientific writing of numbers
```

## Loading and preprocessing the data
```{r}
# unzip and load the data file
df <- read.csv(unz("activity.zip", "activity.csv"))

# the "interval" variable counts 5-minutes intervals. However, this is not very handy and we'd like to replace it by a timestamp:
df <- df %>%
        mutate(interval =  sprintf("%04d", interval)) # add leading 0 if not yet 4

df$time <- paste(substr(as.character(df$interval), 1,2), substr(as.character(df$interval), 3,4), sep = ":")
df$timestamp <- strptime(paste(df$date, df$time), format = "%Y-%m-%d %H:%M")

df <- data.frame(df)


# and as usual, lets have a first look at the data:
summary(df)

```


## What is mean total number of steps taken per day?

```{r}
# calculate the total number of steps taken per day
stepseveryday <- df %>%
                        select(steps, date, interval)%>%
                        group_by(date) %>%
                        summarize(totalsteps = sum(steps))

# make a histogram of the total number of steps taken each day. as there are 61 days in total (since stepseveryday got 61 rows) I go for a binwidth of 61 to show every number in detail.
p <- ggplot(data = stepseveryday, aes(x = totalsteps)) +
        geom_histogram(binwidth = 61)
p + xlab("Total Number of Steps Taken a Day")

# calculate and report the mean and median of the total number of steps taken per day
mean.totalsteps <- mean(stepseveryday$totalsteps, na.rm = TRUE)
median.totalsteps <- median(stepseveryday$totalsteps, na.rm = TRUE)
mean.totalsteps
median.totalsteps

```

The mean of the total number of steps taken per day is `r mean.totalsteps`. The median of the total number of steps taken per day is `r median.totalsteps`.

## What is the average daily activity pattern?
```{r}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
avg.steps.per.interval <- df %>%
                                select(steps, date, interval) %>%
                                group_by(interval) %>%
                                summarize(averagesteps = mean(steps, na.rm = TRUE)) %>%
                                data.frame()

p <- ggplot(data = avg.steps.per.interval, aes(x = as.numeric(interval), y = averagesteps)) +
        geom_line()
p + xlab("Intervals") + 
    ylab("Average Number of Steps")

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
indexofmax <- which(avg.steps.per.interval$averagesteps == max(avg.steps.per.interval$averagesteps))
theinterval <- avg.steps.per.interval[indexofmax,"interval"]
theinterval
```

On average, across all days in the dataset, the maximum number of steps have been counted during the 5-minute Interval from `r theinterval`:00 to `r theinterval`:59.

## Imputing missing values

```{r}
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

# first, lets have a look on how the NAs are spread over the columns
summary(select(df, steps, date, interval)) # its, as expected, in the stepse steps measurements only

total.na <- sum(is.na(df$steps))
total.na
```

The total number of missing values in the dataset is `r total.na`

#### Strategy to Impute missing Values in the Dataset

I will go with a replacement by the rounded mean for that 5-minute interval. 

```{r}
#   Create a new dataset that is equal to the original dataset but with the missing data filled in.
df.NAsimputed <- df %>%
                        select(steps, date, interval)

df.NAsimputed$steps <- mapply(
                function(x,y){
                        if(is.na(x)){
                                index <- which(avg.steps.per.interval$interval == y)
                                x <- round(avg.steps.per.interval[index,"averagesteps"])
                               }
                        x
                        },
                df.NAsimputed$steps,
                df.NAsimputed$interval
                )    

summary(df.NAsimputed) # no NAs left, as desired

#  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

# we need to recalculate the total nr of steps taken every day
stepseveryday.NAsimputed <- df.NAsimputed %>%
                        group_by(date) %>%
                        summarize(totalsteps = sum(steps))

# histplot
p <- ggplot(data = stepseveryday.NAsimputed, aes(x = totalsteps)) +
        geom_histogram(binwidth = 61)
p + xlab("Total Number of Steps Taken a Day") 

# calculate and report the mean and median of the total number of steps taken per day
mean.totalsteps.NAimp <- mean(stepseveryday.NAsimputed$totalsteps)
median.totalsteps.NAimp <- median(stepseveryday.NAsimputed$totalsteps)
mean.totalsteps.NAimp
median.totalsteps.NAimp
```


#### Discussion Differences Dataset original and Dataset with imputed NAs

The mean of the original data set neglecting the NAs was `r mean.totalsteps` and the median `r median.totalsteps`.
After the imputation of the NAs by the rounded mean of this particular interval over all obeserved days, we gt a mean of `r mean.totalsteps.NAimp` which is a difference of `r mean.totalsteps - mean.totalsteps.NAimp` compared to the original data.
The median lowers by `r median.totalsteps - median.totalsteps.NAimp` to `r median.totalsteps.NAimp`.

Also the histograms of the total numbers of steps taken each day differ.
```{r}
comparesteps <- rbind(data.frame(stepseveryday, filter = "original"),
                      data.frame(stepseveryday.NAsimputed, filter = "NAsimputed"))

p <- ggplot(data = comparesteps, aes(x = totalsteps, fill = filter)) +
        geom_histogram(binwidth = 61, position = "dodge")
p + xlab("Total Number of Steps Taken a Day") 

```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
#  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Sys.setlocale("LC_TIME", "English") #as my sys language is not english
df.NAsimputed$date <- as.Date(df.NAsimputed$date)
df.NAsimputed$weekday <- df.NAsimputed$date %>%
                                weekdays() 

df.NAsimputed$daytype <- sapply(df.NAsimputed$weekday,
                                function(x){
                                        if(x %in% c("Saturday", "Sunday")){
                                                x <- "weekend"
                                        } else {
                                                x <- "weekday"        
                                                }
                                        }) %>%
                        as.factor()


#  Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

avg.steps.per.interval.NAimputed <- df.NAsimputed %>%
                                select(steps, date, interval, daytype) %>%
                                group_by(daytype,interval) %>%
                                summarize(averagesteps = mean(steps, na.rm = TRUE)) %>%
                                data.frame()

p <- ggplot(data = avg.steps.per.interval.NAimputed, 
            aes(x = as.numeric(interval), y = averagesteps)) +
        geom_line()
p + facet_grid(daytype ~ .) +
        xlab("Intervals") + 
        ylab("Average Number of Steps")
 

```

We observe that on average the muber of steps are higher during the weekend. Except for the measurements between 5 am and 10 am. During these times, we see that there were almost twice as much steps during the weekdays.